Notes

### 2022-09-22 notes/notes.tex
clarify proof of gradient

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
    \item $\mathrm{vec}: \mathbb{R}^{p \times q} \rightarrow \mathbb{R}^{pq}$ is the vectorisation operator. It can be understood as stacking the columns of a matrix in one long vector. 
    \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
    $$
    \mathrm{diag}\left(
    \begin{bmatrix}
        a_1 \\ 
        \vdots \\
        a_p
    \end{bmatrix}
    \right)
    =
    \begin{bmatrix}
        a_1 & 0 & \dots & 0 \\
        0 & a_2 & \dots & 0 \\
        \vdots & \vdots & \ddots & 0 \\
        0 & 0 & \cdots & a_p
    \end{bmatrix}
    .$$
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
        This is because the Kronecker product has the following convenient property \cite[Section 12.3]{Golub2013}: $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$.
    \item For a vector-valued function $F(x)$, the Jacobian matrix is denoted as $\frac{\partial F}{\partial x}$.
\end{itemize}

@@

Thus, in practice, we are interested in the optimisation problem
$$
\min_{v} f(v) =
\frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
\quad\text{where}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
.$$

Most optimisation algorithms require at least the gradient of $f$. This is given by the following:
\begin{proposition}
    Let $g = [g_1, \dots, g_n]$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
    where $e^v$ denotes elementwise exponentiation and $\odot$ is the Hadamard (elementwise) product.
\end{proposition}
\begin{proof}

Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
Define $$
F(v) := \log h.(\hat{y}(v))
.$$
Since 
$$
\hat{y}(v) = M_{shift} W(v) \tilde{x}
=
(\tilde{x}^T \otimes M_{shift}) \mathrm{vec} W(v)
=
(\tilde{x}^T \otimes M_{shift}) U e^v
,$$
we have
$$
\frac{\partial F}{\partial v}
=
\mathrm{diag}(d \log h.(\hat{y}(v))) \frac{\partial \hat{y}}{\partial v}
=
\underbrace{\mathrm{diag}(d \log h.(\hat{y}(v)))}_H U^T (\tilde{x} \otimes M_{shift}^T) e^{\mathrm{diag}\, v}
.$$
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =
\left(\frac{\partial F}{\partial v}\right)^T F(v)
&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) H \log h.(\hat{y})
\end{align*}
Since $U$ is the unitary map that maps a vector onto (the vectorisation of) the corresponding diagonal matrix, $U^T$ maps the vectorisation of a matrix onto its diagonal. If $\mathrm{Diag}: \mathbb{R}^{p \times p} \rightarrow \mathbb{p}$ maps a matrix to its diagonal, we have 
\begin{align*}
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
&=
e^{\mathrm{diag}\, v} \mathrm{Diag}(M_{shift}^T H \log h.(\hat{y}) \tilde{x}^T) \\
&=
e^{\mathrm{diag}\, v} \left((M_{shift}^T H \log h.(\hat{y})) \odot \tilde{x}\right) \\
&=
e^{v} \odot \left(M_{shift}^T H \log h.(\hat{y})\right) \odot \tilde{x} \\
&=
e^v \odot (M_{shift}^T g(\hat{y})) \odot \tilde{x}
\end{align*}
where $g$ is the elementwise application of $x \mapsto \log h(x) \frac{\mathrm{d}}{\mathrm{d}x}(\log h(x))$. Thus, 
$$
g(\hat{y}_i) = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
$$

The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term is simply $\lambda v$.

\end{proof}

>>>>

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
    \item $\mathrm{vec}: \mathbb{R}^{p \times q} \rightarrow \mathbb{R}^{pq}$ is the vectorisation operator. It can be understood as stacking the columns of a matrix in one long vector. 
    \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
    $$
    \mathrm{diag}\left(
    \begin{bmatrix}
        a_1 \\ 
        \vdots \\
        a_p
    \end{bmatrix}
    \right)
    =
    \begin{bmatrix}
        a_1 & 0 & \dots & 0 \\
        0 & a_2 & \dots & 0 \\
        \vdots & \vdots & \ddots & 0 \\
        0 & 0 & \cdots & a_p
    \end{bmatrix}
    .$$
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
        This is because the Kronecker product satisfies $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$ \cite[eq. 1.3.6]{Golub2013}.
    \item For a vector-valued function $F(x)$, the Jacobian matrix is denoted as $\frac{\partial F}{\partial x}$.
\end{itemize}

@@

Thus, in practice, we are interested in the optimisation problem
\begin{equation}
    \label{eq: cost function parameterised}
\min_{v} f(v)
\quad\text{where}\quad 
f(v)
=
\frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
\quad\text{and}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
.\end{equation}

Most optimisation algorithms require at least the gradient of $f$. This is given by the following:
\begin{proposition}
    Let $g = [g_1, \dots, g_n]$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
    where $e^v$ denotes elementwise exponentiation and $\odot$ is the Hadamard (elementwise) product.
\end{proposition}
\begin{proof}
Let $W = e^{\mathrm{diag}\, v}$ and let $U$ be the matrix corresponding to the linear map $\mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$.
Define $$
F(v) := \log h.(\hat{y}(v))
.$$
By using basic properties of the Kronecker product \cite[eq. 1.3.6]{Golub2013}, we have
$$
\hat{y}(v) = M_{shift} W(v) \tilde{x}
=
(\tilde{x}^T \otimes M_{shift}) \mathrm{vec} W(v)
=
(\tilde{x}^T \otimes M_{shift}) U e^v
,$$
so that
$$
\frac{\partial F}{\partial v}
=
\underbrace{\mathrm{diag}\left(
    \left.\frac{\mathrm{d}}{\mathrm{d}x} \log h(x)\right\vert_{x = \hat{y}(v)}
    \right)}_{:= H}
\frac{\partial \hat{y}}{\partial v}
=
H (\tilde{x}^T \otimes M_{shift}) U e^{\mathrm{diag}\, v}
.$$
The matrix $H$ appearing in this expression is defined by the diagonal elements 
$$
H_{ii} = \frac{1}{h(\hat{y}_i)}\left(\left.\frac{\mathrm{d}}{\mathrm{d}x} h(x)\right\vert_{x = \hat{y}_i} \right)
.$$
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
&=
\left(\frac{\partial F}{\partial v}\right)^T F(v) \\
&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) H \log h.(\hat{y}) \\
&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) g \\
&= e^{\mathrm{diag}\, v} U^T \mathrm{vec}(M_{shift}^T g \tilde{x}^T)
\end{align*}
Since $U$ is the unitary map that maps a vector onto (the vectorisation of) the corresponding diagonal matrix, $U^T$ maps the vectorisation of a matrix onto its diagonal.
Since the diagonal of the rank-1 matrix $M_{shift}^T g \tilde{x}^T$ is $(M_{shift}^T g) \odot \tilde{x}$, the above simplifies to
$$
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
=
e^{\mathrm{diag}\, v} \left( (M_{shift}^T g) \odot \tilde{x} \right) = e^{v} \odot (M_{shift}^T g) \odot \tilde{x}
.$$

The foregoing gives the derivative of the first term in the definition of $f(v)$. The gradient of the second term is simply $\lambda v$.
\end{proof}

With this information, we can optimise \cref{eq: cost function parameterised} using any gradient-based optimisation method. We decided on limited-memory BFGS \cite{Nocedal2006}, which is implimented in SciPy \cite{virtanen2020scipy}.
